Type-Safe Client-Server Communication in Web Applications

### 2014-02-26 paper.tex
Add related work and discussion, restructure introduction.

Development of web applications using conventional means is often no task for
the faint of heart. Code written in Javascript, the only widely supported
language for client-side web applications, is often confusing and error-prone,
much due to the language's lack of modularity, encapsulation facilities and
type safety.

@@

In this paper, we propose a programming model in which web applications are
written as a single entity from which client and server binaries are generated
during compilation. Type annotations in the source program control which parts
are executed on the server and which are executed on the client, and the two
communicate using type safe RPC calls.

Other approaches exist which attempt to tackle the same issue[CITE: STRATIFIEDJS, DUETTO, SUNROOF]
which solve some, but not all, of the associated problems. In particular, our
solution has the following properties:

\begin{itemize}
  \item Type safety: StratifiedJS, being a close relative of the original
        Javascript language, provides type safety for neither RPC calls nor the
        core language and Duetto, being a C++ dialect, is neither memory safe
        nor type safe.
  \item Modularity: our programming model is built as a library on top of
        Haskell and is thus easily portable to any Haskell-to-Javascrit
        compiler. In contrast, StratifiedJS and Duetto are compilers in their
        own right, not usable together with other similar tools.
  \item Purely functional semantics: in contrast to StratifiedJS, Duetto and
        Sunroof, our programming model enables the use of the full Haskell
        language both on the server and the client.
  \item Code reuse: with Sunroof, client code is written in an embedded
        language whereas server code is written in full Haskell, necessitating
        duplicate implementations of functions which may be of use to both the
        client and the server parts. In our solution, once a function is
        written it is usable on both the server and the client.
\end{itemize}

\section{The Haste compiler}
\section{A seamless programming model}
\section{Implementation}
\section{Discussion and related work}

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

@@

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

>>>>

Development of web applications is no task for the faint of heart.
The conventional method involves splitting your program into two logical parts,
writing the one in Javascript, which is notorious even among its proponents for
being wonky and error-prone, and the other in the current language du jour.
Then, two are glued together using whichever home-grown network protocol seems
to fit the application. However, most web applications are conceptually
monolithic entities, making this forced split an undesirable hindrance which
introduces new possibilities for defects, adds development overhead and
prevents code reuse.

Several attempts have been made at tackling this problem, but none is, in our
view, satisfactory. In this paper, we propose a programming model in which a
web application is written as a single program from which client and server
binaries are generated during compilation. Type annotations in the source
program control which parts are executed on the server and which are executed
on the client, and the two communicate using type safe RPC calls. Functions
which are not explicitly declared as server-side or client-side are usable by
either side. Our solution is implemented as a Haskell library requiring minimal
to no compiler support, ensuring portability and modularity.

\paragraph{Motivation}

Code written in Javascript, the only widely supported language for client-side
web applications, is often confusing and error-prone, much due to the
language's lack of modularity, encapsulation facilities and type safety.

@@

\section{A seamless programming model}
\section{Implementation}
\section{The Haste compiler}
\section{Discussion and related work}

Other approaches to this problem exist, notably in the form of the
Conductance\ \cite{conductance}, Duetto\ \cite{duetto} and
Sunroof\ \cite{sunroof} projects, all of which implement some synchronous
abstraction over the network between the client and the server, much like our
solution.

\paragraph{Conductance} Conductance is an application server built on
StratifiedJS, a Javascript language extension which adds a few niceties such as
cooperative multitasking and more concise syntax for many common tasks.
Conductance uses an RPC-based model for client-server communication, much like
our own, but also adds the possibility for the server to independently transmit
data back to the client through the use of shared variables or call back into
the client by way of function objects received via RPC call, as well as the
possibility for both client and server to seamlessly modify variables located
on the opposite end of the network.

While Conductance gets rid of the callback-based programming model endemic to
regular Javascript, it still suffers from many of its usual drawbacks. In
particular, the the weak typing of Javascript poses a problem in that the
programmer is in no way reprimanded by her tools for using server APIs
incorrectly or trying to transmit values which can not be sensibly serialized
and de-serialized, such as DOM nodes. Wrongly typed programs will thus crash, or
even worse, gleefully keep running with erroneous state due to implicit type
conversions, rather than give the programmer some advance warning that something
is amiss.

We are also not completely convinced that the ability to implicitly back and
forth over the network is a unilaterally good thing; while this indeed provides
the programmer some extra convenience, it also requires the programmer to
exercise extra caution to avoid inadvertently sending large amounts of data
back and forth over the network.

\paragraph{Duetto} Duetto is a C++ compiler targeting the web, written from
the ground up to produce code for both client and server simultaneously. It
utilizes the new attributes mechanism introduced in C++11 to designate
functions and data to live on either client or server side. Any calls to a
function on the other side of the network and attempts to access remote data
are implicit, requiring no extra annotations or scaffolding at the call site.

Much like Conductance, Duetto suffers somewhat from its heritage: while the
client-side code is not memory-unsafe, as it is not possible to generate
memory-unsafe Javascript code, its server-side counterpart unfortunately is.
Our reservations expressed about how network communication in Duetto can be
initiated implicitly apply to Duetto as well.

\paragraph{Sunroof} In contrast to the Conductance and Duetto, Sunroof is an
embedded language. Implemented as a Haskell library, it allows the programmer
to use Haskell to write code which is compiled to Javascript and executed on
the client. The language can best be described as having Javascript semantics
with Haskell's type system. Communication between client and server is
accomplished through the use of ``downlinks'' and ``uplinks'', allowing for
data to be sent to and from the client respectively.

Sunroof is completely type-safe, in the DSL itself as well as in the
communication with the Haskell host. However, the fact that client and server
must be written in two separate languages - any code used to generate
Javascript must of course be built solely from the primitives of the Sunroof
language, precluding use of general Haskell code - makes code reuse hard.

\paragraph{Advantages to our approach} We believe that our approach has a
number of distinct advantages to the aforementioned attacks on the problem.

Our approach gives the programmer access to the same strongly typed,
general-purpose functional language on both client and server; any code which
may be of use to both client and server is effortlessly shared, leading to less
duplication of code and increased possibilities for reusing third party
libraries.

Any and all communication between client and server is both strongly typed
and made explicit by the use of the `onServer` function, with the programmer
having complete control over the serialization and de-serialization of data
using the appropriate type classes. Aside from the obvious advantages of type
safety, making the crossing of the network boundary explicit helps the
programmer make an informed decision as to when and where server communication
is appropriate.

Our programming model is implemented as a library, assuming only two Haskell
compilers, one targeting Javascript and one targeting the programmer's server
platform of choice. While we use Haste as our Javascript-targeting compiler,
modifying our implementation to use GHCJS or even the Javascript backend of UHC
would be entirely straightforward. This implementation not only allows for
greater flexibility, but also eliminates any potentially hairy interaction with
compiler internals.

\paragraph{Limitations} Unlike any of the related works, our approach takes a
firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an `onClient`
counterpart to `onServer` would be relatively straightforward.

\section{Future work}

\paragraph{Information flow control} Web applications often make use of a wide
range of third party code for user tracking, advertising, collecition of
statistics and a wide range of other tasks. Any piece of code executing in the
context of a particular web session may not only interact with any other piece
of code executing in the same context, but may also perform basically limitless
communication with third parties and may thus, inadvertently or not, leak
information about the application state. This is obviously highly undesirable
for many applications, which is why there is ongoing work in controlling the
information flow within web applications\ \cite{jsflow}.

While this would indeed provide an effective defence towards attackers and
programming mistakes alike, there is value in being able to tell the two apart,
as well as in catching policy violations resulting from programming mistakes
as early as possible. An interesting venue of research would be to investigate
whether we can take advantage of our strong typing to generate security policies
for such an information flow control scheme, as well as ensure that this policy
is not violated at compile time. This could shorten development cycles as well
as give a reasonable level of confidence that any run time policy violation is
indeed an attempted attack.

\paragraph{Reactive client-server toolkit} While we believe that our work
represents a step forward in making web applications more robust and easier to
develop, the basic model for updating the application's user-visible state is
still essentially imperative. Implementing an FRP-style\ \cite{ppfrp} GUI
programming model on top of our client-server abstraction may prove a fruitful
way to further improve the programmability of web-based distributed
applications.

@@

\bibitem[Bracker, Gill (2014)]{sunroof}
J. Bracker and A. Gill, Sunroof: A Monadic DSL for Generating JavaScript, Practical Aspects of Declarative Languages, 65-80

\bibitem[Conductance application server (2014)]{conductance}
THE INTERNETS, http://conductance.io

\bibitem[Duetto compiler (2014)]{duetto}
THE INTERNETS, http://leaningtech.com/duetto/

\bibitem[Elliott (2009)]{ppfrp}
C. M. Elliott, Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell

\bibitem[Hedin et al (2014)]{jsflow}
D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld, JSFlow: Tracking Information Flow in JavaScript and its APIs
